---
title: "viz_part_2"
author: "Tim Hauser"
output: github_document
---

## Import data

```{r}
library(tidyverse)
library(patchwork)
library(ggridges)
```

Copy in this code from course website to load weather dataset:

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>%
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) %>%
  select(name, id, everything())

weather_df
```

Remember the old scatterplot:

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5)
```

## Labels

Using labs to set title of graph, x, & y axis and add a caption 

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package"
  )
```

## Scales

Using scale_x_continous to manually set and label ticks --> scale_x_* and scale_y_* where * depends on the type of variable mapped to the x and y aesthetics (i.e. continuous vs discrete).

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package") + 
  scale_x_continuous(
    breaks = c(-15, 0, 15), 
    labels = c("-15º C", "0", "15"))
```

Using scale_x/y_* to "log" or "sqrt" transformation of a scale and to set the Y scale to the right:

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package") + 
  scale_x_continuous(
    breaks = c(-15, 0, 15), 
    labels = c("-15ºC", "0", "15"),
    limits = c(-20, 30)) + 
  scale_y_continuous(
    trans = "sqrt", 
    position = "right")
# trans = "log" for log transformation
```

Analogously to scale_x/y_*, there are scales corresponding to other aestheics --> common are used to control the color aesthetic. For example, arguments to scale_color_hue() control the color scale and the name in the plot legend. 

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package") + 
  scale_color_hue(name = "Location", h = c(100, 300))
```

Using vridis color package (you need to make the location name a discrete variable to work) achieves the same result as above and it's much easier, so use this code instead as a standard:

```{r}
ggp_temp_plot = 
  weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package"
  ) + 
  viridis::scale_color_viridis(
    name = "Location", 
    discrete = TRUE
  )
# We used discrete = TRUE because the color aesthetic is mapped to a discrete variable. In other cases (for example, when color mapped to prcp) you can omit this argument to get a continuous color gradient
```

## Themes

Themes are used to modify non-data elements of a plot – they don’t change mappings or how data are render, but control things like background color and location of the the legend. Using themes can help with general plot appearance.

```{r}
ggp_temp_plot + 
  theme(legend.position = "bottom")
```

Minimal theme:

```{r}
ggp_temp_plot + 
  theme_minimal() + 
  theme(legend.position = "bottom")
```

Classic theme:

```{r}
ggp_temp_plot + 
  theme_classic () + 
  theme(legend.position = "bottom")
```

Black & White theme:

```{r}
ggp_temp_plot + 
  theme_bw() + 
  theme(legend.position = "bottom")
```

Excel theme:

```{r}
ggp_temp_plot + 
  ggthemes::theme_excel() + 
  theme(legend.position = "bottom")
```


## Setting Options

Many of these things (i.e., themes and color scheme) are usually set up in advance, i.e., in the first code chunk that is put in the document.This one here is usually used:


library(tidyverse)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d


## Learning Assessment

Revisit the plot showing tmax against date for each location. Use labels, scale options, and theme changes to improve the readability of this plot.

```{r}
ggplot(weather_df, aes(x = date, y = tmax, color = name)) + 
  geom_smooth(se = FALSE) + 
  geom_point(aes(size = prcp), alpha = .75) + 
  labs(
    title = "Temperature plot",
    x = "Date",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package"
  ) + 
  viridis::scale_color_viridis(discrete = TRUE) + 
  theme_minimal() + 
  theme(legend.position = "bottom")
```

## Data is geom()

First create two separate data frames:

```{r}
central_park_df = 
  weather_df %>% 
  filter(name == "CentralPark_NY")

waikiki_df = 
  weather_df %>% 
  filter(name == "Waikiki_HA")

ggplot(waikiki_df, aes(x = date, y = tmax, color = name)) + 
  geom_point() + 
  geom_line(data = central_park_df)
```


## patchwork

We’ve seen facetting as an approach to create the “same plot” for several levels of a categorical variable, and this can get you pretty far. Sometimes, though, you want to show two or three fundamentally different plots in the same graphic: you may want to juxtapose a scatterplot and a boxplot, or show scatterplots illustrating relationships between different variables. In this case, a solution is to create each of the panels you want separately and combine panels using tools in the patchwork package:

```{r}
tmax_tmin_p = 
  weather_df %>% 
  ggplot(aes(x = tmax, y = tmin, color = name)) + 
  geom_point(alpha = .5) +
  theme(legend.position = "none")

prcp_dens_p = 
  weather_df %>% 
  filter(prcp > 0) %>% 
  ggplot(aes(x = prcp, fill = name)) + 
  geom_density(alpha = .5) + 
  theme(legend.position = "none")

tmax_date_p = 
  weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name)) + 
  geom_point(alpha = .5) +
  geom_smooth(se = FALSE) + 
  theme(legend.position = "bottom")

(tmax_tmin_p + prcp_dens_p) / tmax_date_p
```

## Data manipulation

Reorder boxplots by average (not 100% sure...) tmax:

```{r}
weather_df %>%
  mutate(name = fct_reorder(name, tmax)) %>% 
  ggplot(aes(x = name, y = tmax)) + 
  geom_boxplot()
```


From website example; reorders name “by hand”:

```{r}
weather_df %>%
  mutate(name = forcats::fct_relevel(name, c("Waikiki_HA", "CentralPark_NY", "Waterhole_WA"))) %>% 
  ggplot(aes(x = name, y = tmax)) + 
  geom_violin(aes(fill = name), color = "blue", alpha = .5) + 
  theme(legend.position = "bottom")
```

Another website example;reorders name according to tmax values in each name:

```{r}
weather_df %>%
  mutate(name = forcats::fct_reorder(name, tmax)) %>% 
  ggplot(aes(x = name, y = tmax)) + 
  geom_violin(aes(fill = name), color = "blue", alpha = .5) + 
  theme(legend.position = "bottom")
```


## Others

We import the pulse dataset

```{r}
pulse_data = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") %>%
  janitor::clean_names() %>%
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi") %>%
  select(id, visit, everything()) %>%
  mutate(
    visit = recode(visit, "bl" = "00m"),
    visit = factor(visit, levels = str_c(c("00", "01", "06", "12"), "m"))) %>%
  arrange(id, visit)



pulse_df %>% 
  ggplot(aes(x = visit, y = bdi)) +
  geom_boxplot()
```

