---
title: "viz_part_1"
author: "Tim Hauser"
output: github_document
---

## Import data

```{r}
library(tidyverse)
library(ggridges)
```

Copy in this code from course website:

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>%
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) %>%
  select(name, id, everything())
```

Let's make a scatterplot

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) + geom_point()
```

Let's make the a similar  scatterplot, but differently using piping (also added a filter below)

```{r}
weather_df %>%
  drop_na() %>% 
  filter(name == "CentralPark_NY") %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point()
```

Let's keep making the same plot but different

```{r}
weather_scatterplot =
  weather_df %>% 
  drop_na() %>% 
  ggplot(aes(x = tmin, y = tmax))

weather_scatterplot + geom_point()
```

## Let's fancy this up a bit

Adding color + legend, a smooth curve

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name)) +
  geom_smooth()
# here geom_smooth applies to whole
```

Same as above but this time smooth curve applied to each color, then made points transparent and took out SE of smooth curve

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE)
# here geom_smooth applies to each color
```

Make separate panels separated by whatever variable we are interested in

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ name)
# row against columns instead add: facet_grid(. ~ name)
```

`tmax` vs. `tmin` is boring, let's spice it up some

```{r}
weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name)) + 
  geom_point(aes(size = prcp), alpha = .3) +
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)
# made size of dots according to prcp, i.e., added another layer of information
```

## Learning Assignment 1

```{r}
weather_df %>% 
  filter(name == "CentralPark_NY") %>% 
  mutate(
    tmax_fahr = tmax * (9 / 5) + 32,
    tmin_fahr = tmin * (9 / 5) + 32) %>% 
  ggplot(aes(x = tmin_fahr, y = tmax_fahr)) +
  geom_point(alpha = .5) + 
  geom_smooth(method = "lm", se = FALSE)
```

## Odds and ends

```{r}
ggplot(weather_df, aes(x = tmax, y = tmin)) + 
  geom_hex()
# When you’re making a scatterplot with lots of data, there’s a limit to how much you can avoid overplotting using alpha levels and transparency. In these cases geom_hex(), geom_bin2d(), or geom_density2d() can be handy
```


## Unvariate plots...

Histograms, barplots, boxplots, violins, ...

Histogram:

```{r}
weather_df %>% 
  ggplot(aes(x = tmax, fill=name)) + 
  geom_histogram() +
  facet_grid(. ~ name)
# fill= works better with histograms instead of color=
# often it makes sense to put historams in different panels
```

Density plot:

```{r}
weather_df %>% 
  ggplot(aes(x = tmax, fill=name)) + 
  geom_density(alpha=0.3)
```

Boxplots:

```{r}
weather_df %>% 
  ggplot(aes(x = name, y = tmax, fill=name)) + 
  geom_boxplot()
```

Violin plots:

```{r}
weather_df %>% 
  ggplot(aes(x = name, y = tmax, fill=name)) + 
  geom_violin()
```

Ridgeplots:

```{r}
weather_df %>% 
  ggplot(aes(x = tmax, y = name, fill=name)) + 
  geom_density_ridges()
```

## saving and embedding plots

```{r}
weather_scatterplot =
  weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point(aes(size = prcp), alpha = .3) + 
  geom_smooth(se = FALSE) +
  facet_grid(. ~ name)

weather_scatterplot

ggsave(
  file = "results/weather_scatterplot.pdf", 
  plot = weather_scatterplot,
  width = 8, height = 5)
```

```{r, fig.width = 8}
weather_scatterplot
# can tweak how plot is outputted in the markdown file in the code chunk
```


